Evaluating infectious disease forecasts \\ with allocation scoring rules

### 2024-01-31 alloscore_manuscript/alloscore-application-shorter.Rnw
adding some additional discussion comments.

<<thermometer-plot, fig.height=7, fig.width=7.5, fig.cap="Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the ten highest hospitalization counts. For each state, the dark black line shows the data observed when the forecast was made and the grey line shows eventually observed counts. The side-by-side shaded regions show the median (solid horizontal line) and 50%, 80% and 95% prediction intervals for the two selected models. The forecasts were made for new hospitalizations on January 3, 2022 (vertical dashed line, with number of hospitalizations indicated by red horizontal line segment at the intersection of the dashed line and the grey line of data). The vertical bars with red and yellow shading show the allocations: the red bar goes from zero to the level of the observed number of hospitalizations while the yellow bar shows how many resources were suggested by the model to be allocated to that location. The allocations (yellow bar) may be more or less than the observed number, therefore exposed red bar indicates unmet need, while the yellow bar extending above the observed number of hospitalizations means that excess resources were suggested for that given location.">>=
source("R/plot_functions.R")

@@

In epidemiological forecasting, well-known proper scoring rules such as the log score or variations on the continuous ranked probability score (such as the weighted interval score, or WIS) have been frequently utilized to evaluate probabilistic forecasts.
Often models are ranked with respect to accuracy according to a particular score, but without reference to any underlying decision-making process for which that score was designed or from which it might be derivable as a Bayes scoring rule.
With careful thought and collaboration between modelers and public health officials, we argue that scores that are more aligned with public health decisions could be developed to inform specific problems.
We have demonstrated that forecast evaluation methods that are tied to a specific decision making context can yield model rankings that differ substantively from those based on standard measures of forecast skill.
In particular, the allocation score penalizes models that mischaracterize the relative magnitude of resource need across different locations, while WIS penalizes forecasts that do not capture the absolute magnitude of the target quantity in each location.

@@

While our example used a hypothetical scenario to illustrate a potential application of the allocation score, resource allocation decisions are a realistic example of the kinds of decisions that could motivate more targeted forecasting exercises.
One real-world example, which was the motivating example presented in the work above, is allocation of ventilators during a respiratory viral pandemic\citep{huang_stockpiling_2017}.
Other examples include the allocation of a limited stockpile of vaccinations\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests\citep{du_optimal_2022,pasco_covid-19_2023}.

In practice, there are many potential users of forecasts with many different decision making problems.
Not all can be easily modeled as an effort to minimize the expectation of a transparent loss function.
And even those that are easily framed as expected loss minimization may differ enough that no single score would be
appropriate for all users.
Ideally, targeted forecasting tools could be developed through close collaboration between modelers and public health officials.
However, this may only be possible in settings with sufficient staffing on both an analytics and a public health team.
Increasingly, collaborative modeling hubs are being used to generate ``one-size-fits-all'' forecasts for many locations at once.
In these settings, where tailored models are not available, it still could be possible to evaluate contributed models using a set of multiple scores to support public health end users in understanding the value of forecasts as an input to their particular decision making contexts.
% NGR: leaving this out for now as it feels like too much detail?
% This may be tricky to operationalize in the setting of a general forecast hub. It matters how you elicit and represent probabilistic forecasts (quantiles? samples? cdfs?).

@@

In conclusion, we argue that the way modelers and policymakers view and evaluate forecasts should change depending on the specific decision-making context.
Defaulting to standard forecast evaluation metrics can mask the utility (or disutility) of certain forecasts, or lead to forecasts being used in decision making contexts very different from those for which they were intended.
New collaborative work between public health officials and modeling teams is needed to assess the value and relevance of the initial findings presented here, including real-time pilot studies or simulation exercises that could be used to inform further development of new or alternative scoring metrics.
We see this work as an initial overture for what we hope will grow to be a large, collaborative body of work more closely coupling applied epidemiological forecasting with public health decision making.

>>>>

<<thermometer-plot, fig.height=7, fig.width=7.5, fig.cap="Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the ten highest hospitalization counts. For each state, the dark black line shows the data observed when the forecast was made and the grey line shows eventually observed counts. The side-by-side shaded regions show the median (solid red horizontal line) and 50%, 80% and 95% prediction intervals for the two selected models. The forecasts were made for new hospitalizations on January 3, 2022 (vertical dashed line, with number of hospitalizations indicated by red horizontal line segment at the intersection of the dashed line and the grey line of data). The vertical bars with purple, blue and red shading show the allocations. The purple bar goes from zero to the amount of need that was met by the allocation from that model to a specific location. A red bar indicates need that exceeded the resource allocation for a location, and a light blue bar shows suggested resource allocation suggested by that model that were above the need for that location.">>=
source("R/plot_functions.R")

@@

In epidemiological forecasting, well-known proper scoring rules such as the log score or variations on the continuous ranked probability score (such as the weighted interval score, or WIS) have been frequently utilized to evaluate probabilistic forecasts.
Often models are ranked with respect to accuracy according to a particular score, but without reference to any underlying decision-making process for which that score was designed or from which it might be derivable as a Bayes scoring rule.
With careful thought and collaboration between modelers and public health officials, we argue that scores that are more aligned with public health decisions could be developed to inform specific problems.
We have demonstrated that forecast evaluation methods that are tied to a specific decision making context can yield model rankings that differ substantively from those based on standard measures of forecast skill, a result that aligns with findings in other fields \citep{leitch1991economicForecastEval, murphy1993whatisagoodforecast, cenesizoglu2012returnPredictionEconValue}.
In particular, the allocation score penalizes models that mischaracterize the relative magnitude of resource need across different locations, while WIS penalizes forecasts that do not capture the absolute magnitude of the target quantity in each location.

@@

While our example used a hypothetical scenario to illustrate a potential application of the allocation score, resource allocation decisions are a realistic example of the kinds of decisions that could motivate more targeted forecasting exercises.
%One real-world example, which was the motivating example presented in the work above, is allocation of ventilators during a respiratory viral pandemic\citep{huang_stockpiling_2017}.
We used ventilators as a notional example for a resource allocation problem because it (a) has been identified as a scarce resource during outbreaks\citep{huang_stockpiling_2017}, (b) could be allocated across different locations, with logistical considerations depending on the geographical scales, and (c) would likely be correlated (although not perfectly so) with hospitalization data.
However, we acknowledge the limitations of this example in that not everyone who is hospitalized needs a ventilator, and new hospital admissions may not be the best stand in for a need that corresponds to a prevalence of severe cases.
Other possible real-world examples include the allocation of a limited stockpile of vaccinations\citep{araz_geographic_2012,persad_fair_2023} or diagnostic tests\citep{du_optimal_2022,pasco_covid-19_2023}.
While there are limitations in the realism of the pairing of a specific resource and forecasted outcome, we emphasize that the important result from this work is how evaluating forecasts using a fundamentally different kind of score can illuminate previously hidden aspects of value in forecasts.

In practice, there are many potential users of forecasts with many different decision making problems.
While outcomes of some decisions could potentially be specified by loss functions, other uses for forecasts may be less quantifiable, such as when forecasts are used to inform a general sense of situational awareness that may feed in unmeasurable ways into other decisions.
And even those that are easily framed as expected loss minimization may differ enough that no single score would be
appropriate for all users.
Ideally, targeted forecasting tools could be developed through close collaboration between modelers and public health officials.
However, this may only be possible in settings with sufficient staffing on both an analytics and a public health team.
Increasingly, collaborative modeling hubs are being used to generate ``one-size-fits-all'' forecasts for many locations at once.
In these settings, where tailored models are not available, it still could be possible to evaluate contributed models using a set of multiple scores to support public health end users in understanding the value of forecasts as an input to their particular decision making contexts.
% NGR: leaving this out for now as it feels like too much detail?
% This may be tricky to operationalize in the setting of a general forecast hub. It matters how you elicit and represent probabilistic forecasts (quantiles? samples? cdfs?).

@@

In conclusion, we argue that the way modelers and policymakers view and evaluate forecasts should change depending on the specific decision-making context.
Defaulting to standard forecast evaluation metrics can mask the utility (or disutility) of certain forecasts, or lead to forecasts being used in decision making contexts very different from those for which they might offer useful guidance.
New collaborative work between public health officials and modeling teams is needed to assess the value and relevance of the initial findings presented here, including real-time pilot studies or simulation exercises that could be used to inform further development of new or alternative scoring metrics.
We see this work as an initial overture for what we hope will grow to be a large, collaborative body of work more closely coupling applied epidemiological forecasting with public health decision making.

### 2024-01-31 alloscore_manuscript/alloscore-application-shorter.tex
adding some additional discussion comments.

The hospitalization data used for scoring forecasts were downloaded on December 03, 2023.

@@

% latex table generated in R 4.3.2 by xtable 1.8-4 package
% Mon Jan 29 16:46:53 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  percap & 865 &  & 1029 & 366 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{Comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/thermometer-plot-1} \caption[Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the ten highest hospitalization counts]{Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the ten highest hospitalization counts. For each state, the dark black line shows the data observed when the forecast was made and the grey line shows eventually observed counts. The side-by-side shaded regions show the median (solid horizontal line) and 50\%, 80\% and 95\% prediction intervals for the two selected models. The forecasts were made for new hospitalizations on January 3, 2022 (vertical dashed line, with number of hospitalizations indicated by red horizontal line segment at the intersection of the dashed line and the grey line of data). The vertical bars with red and yellow shading show the allocations: the red bar goes from zero to the level of the observed number of hospitalizations while the yellow bar shows how many resources were suggested by the model to be allocated to that location. The allocations (yellow bar) may be more or less than the observed number, therefore exposed red bar indicates unmet need, while the yellow bar extending above the observed number of hospitalizations means that excess resources were suggested for that given location.}\label{fig:thermometer-plot}
\end{figure}

>>>>

The hospitalization data used for scoring forecasts were downloaded on January 30, 2024.

@@

% latex table generated in R 4.3.2 by xtable 1.8-4 package
% Wed Jan 31 15:25:28 2024
\begin{table}[ht]
\centering
\begin{tabular}{lrrrr}
  \hline
Model & AS & MWIS & IAS centered at 15k & IAS uniform \\ 
  \hline
CU-select & 669 & 133 & 774 & 326 \\ 
  percap & 865 &  & 1029 & 366 \\ 
  COVIDhub-ensemble & 873 & 159 & 1067 & 438 \\ 
  USC-SI\_kJalpha & 995 & 91 & 1216 & 1097 \\ 
  JHUAPL-Gecko & 1034 & 164 & 1141 & 418 \\ 
  MUNI-ARIMA & 1084 & 169 & 1248 & 440 \\ 
  COVIDhub-trained\_ensemble & 1089 & 169 & 1271 & 823 \\ 
  COVIDhub-baseline & 1175 & 170 & 1317 & 535 \\ 
  JHUAPL-Bucky & 1358 & 102 & 1566 & 1214 \\ 
  JHUAPL-SLPHospEns & 1540 & 129 & 1604 & 1102 \\ 
  UVA-Ensemble & 2469 & 213 & 2635 & 2494 \\ 
   \hline
\end{tabular}
\caption{Comparison of allocation scores (AS), mean weighted interval scores (MWIS), and two varieties of Integrated Allocation Scores (IAS). All metrics are shown for 10 models that made forecasts of hospital admissions for 2022-01-03. Results are sorted by AS. For all metrics, lower scores indicate better accuracy.} 
\label{tab:multi-k-scores}
\end{table}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{figure}
\includegraphics[width=\maxwidth]{figure/thermometer-plot-1} \caption[Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the ten highest hospitalization counts]{Probabilistic forecasts for new hospital admissions and the inferred resource allocations for COVID-19 on January 3, 2022 for the states with the ten highest hospitalization counts. For each state, the dark black line shows the data observed when the forecast was made and the grey line shows eventually observed counts. The side-by-side shaded regions show the median (solid red horizontal line) and 50\%, 80\% and 95\% prediction intervals for the two selected models. The forecasts were made for new hospitalizations on January 3, 2022 (vertical dashed line, with number of hospitalizations indicated by red horizontal line segment at the intersection of the dashed line and the grey line of data). The vertical bars with purple, blue and red shading show the allocations. The purple bar goes from zero to the amount of need that was met by the allocation from that model to a specific location. A red bar indicates need that exceeded the resource allocation for a location, and a light blue bar shows suggested resource allocation suggested by that model that were above the need for that location.}\label{fig:thermometer-plot}
\end{figure}
